# Bananas App {#appendix-example-bananas-model}

The `bananas` app was born out of a "stuck-in-the-house" COVID-19 project when
one of the authors bought some green bananas at the store and took daily
photographs of each fruit.
Later, the data set was used as part of teaching a workshops. The motivation for the app
is that it follows a workflow that is fairly common in all kinds of data science
projects:

1. Have a question to answer: _Is my banana ripe?_
2. Collect data: _Go to the store, buy bananas, set up a ring light and take pictures every day over 3 weeks._
3. Compile the training data: _Classify colour pixels and calculate the relative proportions, score pictures according to ripeness status._
4. Run exploratory data analysis: _Let's explore and visualize the data set._
5. Train a classification model: _Estimate the the banana ripeness class and probability given the colour composition._
6. Build a "scoring engine": _Given some colour inputs for a new fruit, tell me the probability for the ripeness classes._
6. Build a user interface: _Let a non technical user to do the data exploration and classification as part of a web application._

The Bananas data set tracks the ripening colour composition of banana fruits daily over a
3-week period. The full data set can be found in the [GitHub repository](https://github.com/psolymos/bananas)
and R package `bananas` (`install.packages("bananas", repos = "https://psolymos.r-universe.dev")`).
The subset used in the book and the Shiny app constitutes the 6 fruits that were kept at room temperature.

```{r setup_appA, include=FALSE}
options(scipen = 99, digits = 3)
```

The table has the following fields:

- `fruit`: the identifier of the fruit,
- `day`: number between 0 and 20, the number of days since the first set of photographs,
- `ripeness`: the ripeness class of the fruit based on Péter's personal judgement (Under, Ripe, Very, Over),
- `green`, `yellow`, `brown`: colour composition, these 3 values add up to 1 (100%).

```{r eval=FALSE,echo=FALSE}
bananas <- read.csv("data/bananas.csv")
str(bananas)
```

```{r eval=TRUE,echo=FALSE, results='asis'}
if(is_latex_output()){
txt <- '
The colour composition was determined based on colour mapping the pixel values of 
the banana fruits and converting the pixel based 2-dimensional area to proportions
(Table \\@ref(tab:bananas-table)).
'
} else {
txt <- '
The colour composition was determined based on colour mapping the pixel values of 
the banana fruits and converting the pixel based 2-dimensional area to proportions.

The following summary presents the ripeness and the percentage values of 
green, yellow, brown colours.
'
}
cat(txt)
```

```{r bananas-table, eval=TRUE, echo=FALSE}
x <- read.csv("data/bananas.csv")
if (!is_latex_output()) {
    reactable(x,
        defaultPageSize = 10,
        showPageSizeOptions = TRUE,
        pageSizeOptions = c(10, 30, 60, 120),
        highlight = TRUE,
        searchable = TRUE,
        striped = TRUE)
} else {
    capt <- "A summary of the bananas data set presented as fruits (columns) over the days (rows), each cell showing the ripeness (U = Under, R = Ripe, V = Very, O = Over) and the percentage values of green, yellow, brown colours, respectively."
    z <- matrix("", 20, 6)
    rownames(z) <- sort(unique(x$day))
    colnames(z) <- sort(unique(x$fruit))
    for (i in sort(unique(x$day))) {
        for (j in sort(unique(x$fruit))) {
            ij <- x$day == i & x$fruit == j
            # v <- paste0(x$ripeness[ij], " [G=", round(100*x$green[ij]), 
            #     "|Y=", round(100*x$yellow[ij]), "|B=", round(100*x$brown[ij]), "]")
            v <- paste0(substr(x$ripeness[ij], 1, 1), " (", round(100*x$green[ij]), 
                ", ", round(100*x$yellow[ij]), ", ", round(100*x$brown[ij]), ")")
            z[as.character(i),j] <- v
        }
    }
    rownames(z) <- paste0("Day ", sort(unique(x$day)))
    # https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf
    # x[x$fruit == "fruit_14",] |> 
    z |>
    kbl(booktabs = TRUE,
        longtable = TRUE,
        caption = capt) |>
    kable_styling(
        font_size = 8,
        latex_options = c("repeat_header"))
}
```

Figure \@ref(fig:bananas-time) shows the change in colour composition over
the 3 weeks of the experiment. You can see that the proportion of green colour
went down, parallel to that the yellow colour proportion peaked around day 5.
Yellow started decreasing after that while the proportion of brown started
increasing.

We can also present the same information according to the ripeness classes
(Fig. \@ref(fig:bananas-classes)). You can see that the under-ripe
class is characterized by high green proportion and the absence of brown.
The ripe class is characterized by the highest proportion of yellow.
Very ripe bananas have higher proportion of brown while yellow colour is still
the most common. Over ripe bananas are mostly brown.

```{r bananas-time, eval=TRUE, echo=FALSE, out.width="100%", fig.pos = "bt", fig.cap="Colour composition of the bananas over time."}
x <- read.csv("data/bananas.csv")
x$ripeness <- factor(x$ripeness, c("Under", "Ripe", "Very", "Over"))

ggplot(
    data = data.frame(
        prop = c(x$green, x$yellow, x$brown),
        Colours = rep(c("green", "yellow", "brown"), each = nrow(x)),
        day = x$day),
    mapping = aes(
        x = jitter(day, 0.5),
        y = prop,
        group = Colours,
        color = Colours)) +
    geom_point(alpha = 0.75) +
    geom_smooth(
        method = 'loess', 
        formula = 'y ~ x',
        se = FALSE) +
    scale_color_manual(
        values = c(
            green = "#576a26", 
            yellow = "#eece5a", 
            brown = "#5c361f")) +
    ylim(0, 1) +
    xlab("Days") +
    ylab("Proportion") +
    # theme_minimal()
    # theme_classic()
    theme_linedraw()
```

```{r bananas-classes, eval=TRUE, echo=FALSE, out.width="100%", fig.pos = "bt", fig.cap="Colour composition of the bananas by ripeness class."}
ggplot(
    data = data.frame(
        prop = c(x$green, x$yellow, x$brown),
        Colours = factor(
            rep(c("green", "yellow", "brown"), each = nrow(x)),
            c("green", "yellow", "brown")),
        ripeness = x$ripeness),
    mapping = aes(
        x = ripeness,
        y = prop,
        # group = Colours,
        fill = Colours)) +
    geom_boxplot() +
    # coord_flip() +
    # facet_wrap(vars(Colours)) +
    scale_fill_manual(
        values = c(
            green = "#576a26", 
            yellow = "#eece5a", 
            brown = "#5c361f")) +
    xlab("Ripeness") +
    ylim(0, 1) +
    ylab("Proportion") +
    # theme_minimal()
    # theme_classic()
    theme_linedraw()
```

For model training, we chose Support Vector Machines (SVM) to model a multi-level response variable
(Under, Ripe, Very, Over) as a function of the green, yellow, and brown colours.

We used the `e1071` package [@R-e107] in R, and the SVM model's prediction accuracy was 90.8%.
We saved the trained model object as an R binary `.rds` file:

```R
library(e1071)

# Read the bananas data
x <- read.csv("bananas.csv")
x$ripeness <- factor(x$ripeness, c("Under", "Ripe", "Very", "Over"))

# Multinomial classification with Support Vector Machines
m <- svm(ripeness ~ green + yellow + brown,
  data = x,
  probability = TRUE
)

# Two-way table to test prediction accuracy
table(x$ripeness, predict(m))
sum(diag(table(x$ripeness, predict(m)))) / nrow(x)

# Predict ripeness class
predict(m, data.frame(green = 1, yellow = 0, brown = 0), 
  probability = TRUE)
predict(m, data.frame(green = 0, yellow = 1, brown = 0), 
  probability = TRUE)
predict(m, data.frame(green = 0, yellow = 0, brown = 1), 
  probability = TRUE)
predict(m, data.frame(green = 0.1, yellow = 0.2, brown = 0.7), 
  probability = TRUE)

# Save the model object
saveRDS(m, "bananas-svm.rds")
```

We can fit a similar SVM model in Python using scikit-learn (`sklearn`) [@scikit-learn]:

```python
import pandas as pd
from joblib import dump
from sklearn import svm

# Global
x = pd.read_csv('bananas.csv')

# Train SVM
x.loc[x.ripeness == 'Under', 'target'] = 0
x.loc[x.ripeness == 'Ripe', 'target'] = 1
x.loc[x.ripeness == 'Very', 'target'] = 2
x.loc[x.ripeness == 'Over', 'target'] = 3
data_X = x[['green', 'yellow', 'brown']].to_numpy()
data_y = x.target.values
svm_model = svm.SVC(probability = True)
svm_model.fit(data_X, data_y)

#' Predict ripeness class
svm_model.predict_proba([[1, 0, 0]])
svm_model.predict_proba([[0, 1, 0]])
svm_model.predict_proba([[0, 0, 1]])
svm_model.predict_proba([[0.1, 0.2, 0.7]])

# Write model object to file
dump(svm_model, 'bananas-svm.joblib')
```
